Boycotts require significant deterioration in a company's public image to have an impact on their stock prices.

### 2024-04-26 paper/paper.qmd
updates to paper and bibliography

---
title: "Boycotts require significant deterioration in a company's public image to have an impact on their stock prices."
subtitle: "An analysis of the stocks of 3 large U.S. companies boycotted between 1992 to 2024."
author: 
  - Jessica Im
thanks: "Code and data are available at: [https://github.com/jimessica/boycott)](https://github.com/jimessica/boycott)"
date: today
date-format: long
abstract: "Introduction to the topic and its importance. What I did in my research and analysis. What was found from my research and analysis. Significance of my findings."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

## Import cleaned data and turn "Date" into class Date
# Nike
cleaned_data_NKE <- read.csv("../data/analysis_data/cleaned_data_NKE.csv")
cleaned_data_NKE <- 
  cleaned_data_NKE |>
  mutate(Date = lubridate::ymd(Date))
cleaned_data_NKE$Date <- 
  as.Date(cleaned_data_NKE$Date, format='%Y-%m-%d')

# United Parks and Resorts
cleaned_data_PRKS <- read.csv("../data/analysis_data/cleaned_data_PRKS.csv")
cleaned_data_PRKS <- 
  cleaned_data_PRKS |>
  mutate(Date = lubridate::ymd(Date))
cleaned_data_PRKS$Date <- 
  as.Date(cleaned_data_PRKS$Date, format='%Y-%m-%d')

@@

The objective of this paper is to add to existing literature by determining the impact of boycotting on stock prices and the factors involved in its effectiveness, particularly media coverage. The estimand is the change in stock price during the boycott period. The beginning of the boycott is determined by an event which sparks the most media coverage or is mentioned the most in the media. The end of the boycott is determined by mention in the media of a “turning point” for the company where the reason for the boycott is addressed. Since Starbucks is an ongoing boycott at the time of this research, the “end date” is the date research began. The stock prices when the market closed were arbitrarily selected instead of the stock price at open. To investigate, Nike, United Parks and Resorts, and Starbucks are analyzed as large U.S. companies that were famously boycotted during the desired time period. The analysis includes a detailed view of the difference in stock prices overall, during periods of heavy media coverage, and in comparison with market trends. From analysis, the stock prices during the boycott of Nike were found to not be impacted, United Parks and Resorts was heavily impacted, and Starbucks were mildly impacted. The range in impact can be attributed to overall company image, product type, earnings calls, unionization, and social media. This result supports the importance of media coverage, but also the lack of financial repercussions to many boycotted companies.

@@

## United Parks and Resorts

Blackfish is a film focused on the cruel captivity of orcas by focusing on the infamous killer whale, Tilikum, who was kept in captivity at SeaWorld and was involved in the death of 3 people [@WAP]. The film was released in 2013 and shown on major channels like CNN, Netflix, and Universal Studios around the world and over its 14-week box office period in North America, earned \$2,073,582 [@wikiBlackfish]. In March of 2014, California Assemblyman Richard Bloom proposed that orcas are banned from being held captive for entertainment [@CNNorcaban]. Later that year, U.S. Congressmen Adam Schiff and Jared Huffman amended the Agriculture Appropriations Act to require the USDA’s Animal & Plant Health Inspection Service to update and finalize regulations on marine mammal captivity [@SchiffandHuffman]. On behalf of SeaWorld investors, the Rosen Law Firm filed a class action lawsuit against SeaWorld in 2014 for

## Starbucks

On October 9th the Starbucks Workers United posted to their Twitter “Solidarity with Palestine” [@apnews]. This tweet was up for 40 minutes before it was taken down and was originally posted without the approval of union leaders according to Workers United. Even after the post was taken down, posts and retweets from local Starbucks United twitter accounts in support of Palestine and condemning Israel were circulating. On October 13th, Starbucks sent a letter to Workers United, demanding the union to stop using the Starbucks name and similar logo. Workers United retaliated saying that Starbucks was using the tragedy in the Middle East to support their anti-union campaign.

@@

A sense of the dataset should be communicated to the reader. The broader context of the dataset should be discussed. All variables should be thoroughly examined and explained. Explain if there were similar datasets that could have been used and why they were not. If variables were constructed then this should be mentioned, and high-level cleaning aspects of note should be mentioned, but this section should focus on the destination, not the journey. It is important to understand what the variables look like by including graphs, and possibly tables, of all observations, along with discussion of those graphs and the other features of these data. Summary statistics should also be included, and well as any relationships between the variables. If this becomes too detailed, then appendices could be used. Basically, for every variable in your dataset that is of interest to your paper there needs to be graphs and explanation and maybe tables.

## Source Data and Methodology

The source data was obtained from Yahoo Finance Canada [@yahoo] and the R package [@R-tidyquant]. The Yahoo Finance Canada historical data was filtered for the desired date range and daily data before being downloaded. The original data had “Date”, “Open”, “High”, “Low”, “Close”, “Adj Close”, and “Volume”. The only variables of interest were “Date” and “Close” which represented the date of the stock’s price and the price when the market closed respectively.

## Limitations

@@

dataALL <- data.frame(
  "Companies" = c("Nike (NKE)", "United Parks and Resorts (PRKS)", "Starbucks (SBUX)"),
  "Days Observed" = c(N_NKE, N_PRKS, N_SBUX),
  "Date Start" = c(Day1_NKE, Day1_PRKS, Day1_SBUX),
  "Date End" = c(Day2_NKE, Day2_PRKS, Day2_SBUX)
)

@@

## Change in Stock Price
The change in stock price over the length of the boycott is observed in @fig-linegraphNKE, @fig-linegraphPRKS, and @fig-linegraphSBUX. These three graphs show a week before the boycott started to the end of the boycott. @fig-linegraphNKE shows how Nike’s stock price was relatively unstable in the first three years of the boycott, with stock prices rising later in 1991 before dropping slightly and then peaking early in 1993 before descending to a low a year later. A progressive cimb from that point to February of 1997 sees a growth of around 750%. @fig-linegraphPRKS shows that United Parks and Resorts had a steep drop in stock price when ‘Blackfish’ released and had a year of poor stability after before dropping drastically. Afterwards the stock price continues to remain unstable.  @fig-linegraphSBUX graphs the stock price of Starbucks and depicts a steep climb in price in November of 2023 with a peak in early December before a steady climb down over the next 5 months.

@@

```{r}
#| label: fig-linegraphPRKS
#| fig-cap: Changes in United Parks and Resorts Stock Price from One Week Before Blackfish was Released to One Week After Their Pledge to End Orca Breeding
#| echo: false

@@

@tbl-starttoendNKE shows Nike had a small decrease in stock price from a week before the beginning of the boycott. There is also an increase in stock price from the day the boycott started to the day it ended. Relative to the stock price, the change is very large as it increased by 335.82%. @tbl-starttoendPRKS shows that there was a significant decrease of 55.3% in United Parks and Resorts's stock price from the beginning to end of the boycott. Starbucks was relatively unchanged from the start to the end of the boycott with a decrease of only 2.5% as shown in @tbl-starttoendSBUX.

@@

```{r}
#| label: tbl-starttoendPRKS
#| tbl-cap: Changes in United Parks & Resorts Stock Price from One Week Before Blackfish was Released, to the Date it was Released, to their Announcement that they will end their Orca Breeding Program
#| echo: false

@@

```{r}
#| label: fig-linegraphPRKSfocused
#| fig-cap: Changes in United Parks and Resorts During the Month SeaWorld First Announced they Were Impacted by the Release of 'Blackfish'
#| echo: false

@@

```{r}
#| label: tbl-twoweeksSBUX
#| tbl-cap: Changes in Starbucks Stock Price Two Weeks From the Lawsuit
#| echo: false

@@

## Comparing with the Market
@fig-djiNKE compares the stock price of Nike and the Dow Jones Index (DJI) [@dowjones]. Dow Jones Index includes Nike among around 30 other companies, and it is clear that they have very similar trajectories throughout the boycott. The only apparent difference is the bump in Nike’s price between 1996 and 1998. Otherwise, both stocks have a steady upward growth.

@@

```{r}
#| label: fig-spxPRKS
#| fig-cap: Changes in United Parks and Resorts Stock Price in Comparison with the S&P 500 index
#| echo: false

@@

## Nike: Seemingly unaffected
Nike is the only company with an increase in stock price over the boycott period and it is quite significant at 334%. Despite this overall growth, Nike does not remain unaffected by the boycotts. In 1997 the company suffered a significant decrease in comparison to the overall growth in the market as seen in @graph-djiNKE. The steep drop in February

## SeaWorld: The Tainted Consumer Experience

@@

## Weaknesses and next steps

Weaknesses and next steps should also be included.

>>>>

---
title: "Boycotts require significant deterioration in a company's public image to have an impact on their stock prices."
subtitle: "An analysis of the stocks of 3 large U.S. companies boycotted between 1992 to 2024."
author: 
  - Jessica Im
thanks: "Code and data are available at: [https://github.com/jimessica/boycott)](https://github.com/jimessica/boycott)"
date: today
date-format: long
abstract: "Boycotting is the act of withholding supporting for an organization to push for change due to an ethical issue by threatening the organization's finances and integrity. To investigate the impact of boycotting on the stock prices of large U.S. companies, the 1990s anti-sweatshop Nike boycott, the post-Blackfish SeaWorld boycott, and the recent pro-Palestine Starbucks boycott were analyzed. The investigation on the impact of the boycotts on their stock prices revealed that the impact is not always financially significant, but the importance of widespread media coverage is unrivaled."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

## Import cleaned data and turn "Date" into class Date (using parquet file was causing multiple issues)
# Nike
cleaned_data_NKE <- read.csv("../data/analysis_data/cleaned_data_NKE.csv")
cleaned_data_NKE <- 
  cleaned_data_NKE |>
  mutate(Date = lubridate::ymd(Date))
cleaned_data_NKE$Date <- 
  as.Date(cleaned_data_NKE$Date, format='%Y-%m-%d')

# SeaWorld
cleaned_data_PRKS <- read.csv("../data/analysis_data/cleaned_data_PRKS.csv")
cleaned_data_PRKS <- 
  cleaned_data_PRKS |>
  mutate(Date = lubridate::ymd(Date))
cleaned_data_PRKS$Date <- 
  as.Date(cleaned_data_PRKS$Date, format='%Y-%m-%d')

@@

The objective of this paper is to add to existing literature by determining the impact of boycotting on stock prices and the factors involved in its effectiveness, particularly media coverage. The estimand is the change in stock price during the boycott period. The beginning of the boycott is determined by an event which sparks the most media coverage or is mentioned the most in the media. The end of the boycott is determined by mention in the media of a “turning point” for the company where the reason for the boycott is addressed. Since Starbucks is an ongoing boycott at the time of this research, the “end date” is the date research began. The stock prices when the market closed were arbitrarily selected instead of the stock price at open. To investigate, Nike, SeaWorld, and Starbucks are analyzed as large U.S. companies that were famously boycotted during the desired time period. The analysis includes a detailed view of the difference in stock prices overall, during periods of heavy media coverage, and in comparison with market trends. From analysis, the stock prices during the boycott of Nike were found to not be impacted, SeaWorld was heavily impacted, and Starbucks were mildly impacted. The range in impact can be attributed to overall company image, product type, earnings calls, unionization, and social media. This result supports the importance of media coverage, but also the lack of financial repercussions to many boycotted companies.

@@

## SeaWorld (PRKS)

"Blackfish" is a film focused on the cruel captivity of orcas by focusing on the infamous killer whale, Tilikum, who was kept in captivity at SeaWorld and was involved in the death of 3 people [@WAP]. The film was released in 2013 and shown on major channels like CNN, Netflix, and Universal Studios around the world and over its 14-week box office period in North America, earned $2,073,582 [@wikiBlackfish]. In March of 2014, California Assemblyman Richard Bloom proposed that orcas are banned from being held captive for entertainment [@CNNorcaban]. Later that year, U.S. Congressmen Adam Schiff and Jared Huffman amended the Agriculture Appropriations Act to require the USDA’s Animal & Plant Health Inspection Service to update and finalize regulations on marine mammal captivity [@SchiffandHuffman]. On behalf of SeaWorld investors, the Rosen Law Firm filed a class-action lawsuit against SeaWorld in 2014 because the company and some leaders made untrue and misleading statements regarding the impact of the Film on their park attendance and image. SeaWorld renamed itself to United Parks and Recreation, also changing its ticker symbol to PRKS in February of 2024 [@namechange]. For the purpose of this paper, the company will be referred to as SeaWorld.

## Starbucks (SBUX)
On October 9th the Starbucks Workers United (SWU) posted to their Twitter “Solidarity with Palestine” [@apnews]. This tweet was up for 40 minutes before it was taken down and was originally posted without the approval of union leaders according to Workers United. Even after the post was taken down, posts and retweets from local Starbucks United twitter accounts in support of Palestine and condemning Israel were circulating. On October 13th, Starbucks sent a letter to Workers United, demanding the union to stop using the Starbucks name and similar logo. Workers United retaliated saying that Starbucks was using the tragedy in the Middle East to support their anti-union campaign. On October 18th, Starbucks filed a lawsuit against SWU for trademark infringement and SWU filed a countersuit in return claiming there was no violation of trademark laws and that Starbucks defamed Workers United [@newyorkpost]. Use of the TikTok hashtag #boycottstarbucks peaked in November and videos on TikTok like a barista showing the public how to make their most popular drinks have reached millions of views [@nbcstarbucks]. The boycott is now supported by the BDS movement, but was never on their target list [@timenews]. Since the boycott is ongoing, the arbitrary “end date” of the boycott for this stock is March 28th, 2024.

@@

## Source Data and Methodology
The source data was obtained from Yahoo Finance Canada [@yahoo] and the R package [@R-tidyquant]. The Yahoo Finance Canada historical data was filtered for the desired date range and daily data before being downloaded. The original data had “Date”, “Open”, “High”, “Low”, “Close”, “Adj Close”, and “Volume”. The only variables of interest were “Date” and “Close” which represented the date of the stock’s price and the price when the market closed respectively. Three index funds were used as a representaiton of the general market trends. Different index funds were used due to constraints on what information is avilable during what time period, but the S&P 500, the Don Jones Index, and the Nasdaq 100 are all stocks known to arguably represent the market.

@@

dataALL <- data.frame(
  "Companies" = c("Nike (NKE)", "SeaWorld (PRKS)", "Starbucks (SBUX)"),
  "Days Observed" = c(N_NKE, N_PRKS, N_SBUX),
  "Date Start" = c(Day1_NKE, Day1_PRKS, Day1_SBUX),
  "Date End" = c(Day2_NKE, Day2_PRKS, Day2_SBUX)
)

@@

## Change in Stock Price
The change in stock price over the length of the boycott is observed in @fig-linegraphNKE, @fig-linegraphPRKS, and @fig-linegraphSBUX. These three graphs show a week before the boycott started to the end of the boycott. @fig-linegraphNKE shows how Nike’s stock price was relatively unstable in the first three years of the boycott, with stock prices rising later in 1991 before dropping slightly and then peaking early in 1993 before descending to a low a year later. A progressive cimb from that point to February of 1997 sees a growth of around 750%. @fig-linegraphPRKS shows that SeaWorld had a steep drop in stock price when "Blackfish" released and had a year of poor stability after before dropping drastically. Afterwards the stock price continues to remain unstable.  @fig-linegraphSBUX graphs the stock price of Starbucks and depicts a steep climb in price in November of 2023 with a peak in early December before a steady climb down over the next 5 months.

@@

```{r}
#| label: fig-linegraphPRKS
#| fig-cap: Changes in SeaWorld Stock Price from One Week Before "Blackfish" was Released to One Week After Their Pledge to End Orca Breeding
#| echo: false

@@

@tbl-starttoendNKE shows Nike had a small decrease in stock price from a week before the beginning of the boycott. There is also an increase in stock price from the day the boycott started to the day it ended. Relative to the stock price, the change is very large as it increased by 335.82%. @tbl-starttoendPRKS shows that there was a significant decrease of 55.3% in SeaWorld's stock price from the beginning to end of the boycott. Starbucks was relatively unchanged from the start to the end of the boycott with a decrease of only 2.5% as shown in @tbl-starttoendSBUX.

@@

```{r}
#| label: tbl-starttoendPRKS
#| tbl-cap: Changes in United Parks & Resorts Stock Price from One Week Before "Blackfish" was Released, to the Date it was Released, to their Announcement that they will end their Orca Breeding Program
#| echo: false

@@

```{r}
#| label: fig-linegraphPRKSfocused
#| fig-cap: Changes in SeaWorld During the Month SeaWorld First Announced they Were Impacted by the Release of "Blackfish"
#| echo: false

@@

```{r}
#| label: tbl-twoweeksSBUX
#| tbl-cap: Changes in Starbucks Stock Price Two Weeks From the Lawsuit on October 18th, 2023
#| echo: false

@@

## Comparing with the Market
@fig-djiNKE compares the stock price of Nike and the Dow Jones Index (DJI). Dow Jones Index includes Nike among around 30 other companies, and it is clear that they have very similar trajectories throughout the boycott. The only apparent difference is the bump in Nike’s price between 1996 and 1998. Otherwise, both stocks have a steady upward growth.

@@

```{r}
#| label: fig-spxPRKS
#| fig-cap: Changes in SeaWorld Stock Price in Comparison with the S&P 500 index
#| echo: false

@@

## Nike: Seemingly unaffected
Nike is the only company with an increase in stock price over the boycott period and it is quite significant at 334%. Despite this overall growth, Nike does not remain unaffected by the boycotts. In 1997 the company suffered a significant decrease in comparison to the overall growth in the market as seen in @fig-djiNKE. The steep drop in February

## SeaWorld: The Tainted Consumer Experience
SeaWorld was the most affected by the boycott of the three companies examined with a decrease in overall stock price of 55.3%. On August 13th, 2014 SeaWorld announced that their declining attendance was partially a result of negative backlash after the release of the film “Blackfish” [@SECvSeaWorld]. The impact of this reveal is seen in @fig-linegraphPRKSfocused where there is a drastic drop in stock price at this time. The relative unaffected stock price before this may be attributed to the withholding of information by the company. In 2018, SeaWorld and their former Chief Executive Officer were sued for making untrue or misleading statements regarding the impact of “Blackfish” on the company’s business and reputation [@SECvSeaWorld]. There was also a class-action suit against the former Vice President of Communications for the same act. They were ordered to pay $5,099,338 in disgorgement, prejudgment interest, and civil penalties to any investors who purchased stock between December 20th, 2013 and August 13th, 2014. While there was a lot of media coverage and boycotting of SeaWorld following the release of “BlackFish”, the stock likely remained relatively stable due to the company and particular leadership misleading investors. Thus, though the impact was delayed, the boycott had a severe negative effect on SeaWorld’s stock. Unlike Nike and Starbucks, SeaWorld’s boycott was the response to a very popular and sensitive movie whose subjects would be directly seen at the park if a patron were to visit. The large impact on SeaWorld’s overall stock price during this time is likely influenced by the fact that visitors would be paying to see the abused animals mentioned in the film, while consumers at Nike and Starbucks are not directly seeing the company’s actions by purchasing their products. In addition, while SeaWorld was a fun and popular park to visit, it did not have nearly as many regular customers as businesses like Nike and Starbucks. SeaWorld’s boycott also received a lot of media attention from the film's box-office success and shocking revelation to an experience very popular among families. Additionally, the scandal of the company misleading investors worsened the company’s overall reliability even more. Overall, many factors led to the significant fall of SeaWorld’s stock price, but the impact of heavy media attention which continues today, likely played a very large role.

@@

## Next Steps

The impact of boycotts is still up for debate, but it is clear that without relentless media attention impacting the company's image to the general public, boycotts will have little impact on the stock prices of large U.S. companies. As boycotts continue to rise due to the overbearing power of fewer companies, it is important for activists to consider what priorities will help make a financial difference to these companies and invoke more drastic change. Investors are rarely impacted by just the existence of a boycott without evidence of it affecting the bottom-line, as shown by the delayed fall of SeaWorld's stock. If the
